---
title: "Tutorial 5 Instrumental Variables"
pagetitle: Tutorial_5
---

Instrumental variable estimation is a crucial tool in causal inference, allowing researchers to address endogeneity and obtain unbiased estimates when treatment assignment is not purely random. This tutorial will guide you through implementing instrumental variable estimation in R using Two-Stage Least Squares (2SLS), as well as testing for endogeneity, weak instruments, and overidentification.

By the end of this tutorial, you will be familiar with:

1. 2SLS in R

2. Assessing instrument validity

# Front-end Matters

We use `AER` package for 2SLS.

```{r}
#install.packages("AER")
library(AER)
library(stargazer)
```

# Introducing dataset

We use the `CigarettesSW` data from the `AER` package. It is a panel data set that contains observations on cigarette consumption and several economic indicators for all 48 continental federal states of the U.S. from 1985 to 1995. We consider data for the cross section of states in 1995 only.

We want to know the relationship between **after-tax average real price per pack of cigarettes** (x) and **the number of cigarette packs per capita sold** (y). Since there is simultaneous causality between demand and supply, we need to use instrumental variable regression.

The The instrumental variable we are going to use for instrumenting the endogenous regressor is *SalesTax*, the portion of taxes on cigarettes arising from the general sales tax. The idea is that *SalesTax* is a relevant instrument as it is included in the after-tax average price per pack. Also, it is plausible that *SalesTax* is exogenous since the sales tax does not influence quantity sold directly but indirectly through the price.

```{r}
data("CigarettesSW")
summary(CigarettesSW)

# compute real per capita prices
CigarettesSW$rprice <- with(CigarettesSW, price / cpi)

#  compute the sales tax
CigarettesSW$salestax <- with(CigarettesSW, (taxs - tax) / cpi)

# generate a subset for the year 1995
c1995 <- subset(CigarettesSW, year == "1995")
```

# 2SLS

## Simple regression model

We first need to check if the independent variable (price) and the instrumental variable (sale tax) is correlated.

```{r}
cor(CigarettesSW$salestax, CigarettesSW$rprice)
```

Since x and z are highly correlated, we can move on to the 2SLS estimation.

**Step 1**, $$log(Price)=\delta_0+\delta_1Tax+v$$

```{r}
cig_s1 <- lm(log(rprice) ~ salestax, data = c1995)
summary(cig_s1)
```

**Step 2**, we estimate the fitted values obtained by the first stage regression

```{r}
lcigp_pred <- cig_s1$fitted.values
```

**Step 3**, we run the second stage regression $$log(packs)=\beta_0+\beta_1\hat{log_price}+\mu$$

```{r}
cig_s2 <- lm(log(c1995$packs) ~ lcigp_pred)
summary(cig_s2)
```

Now we have the model $$log(pack)=9.72-1.08log(price)$$

Comparing the results of OLS regression, first, and second stage.

```{r}
cig_ols <- lm(log(packs) ~ log(rprice),data=c1995)

stargazer(cig_ols,cig_s1,cig_s2,type = "text")
```

Using OLS regression, a 1% increase in log(price) is associated with 1.21% decrease in log(pack). Using the 2SLS estimation, 1% increase in log(price) is associated with 1.08% decrease in log(pack), which is lower effect.

However, although 2SLS correctly estimates the coefficient, the standard error needs to be corrected. The function `ivreg()` from the package `AER` carries out 2SLS procedure automatically and generates the same coefficient and corrected standard error. Now we see the standard error of *log(rprice)* in 2SLS model (0.317) is higher than the standard error from the OLS regression (0.216). The 2SLS model also reports a lower level of significance.

```{r}
cig_ivreg <- ivreg(log(packs) ~ log(rprice) | salestax, data = c1995)
summary(cig_ivreg)
```

## Multiple regression model

After learning the simple regression, let's add *income* as a control variable.

```{r}
# add rincome to the dataset
CigarettesSW$rincome <- with(CigarettesSW, income / population / cpi)
CigarettesSW$cigtax <- with(CigarettesSW, tax/cpi)

c1995 <- subset(CigarettesSW, year == "1995")
```

We first estimate the model with one instrument (sales tax) and one control (income). In the function below, left of `|` shows the dependent variable *log(packs)*, the endogenous variable *log(rprice)*, and the exogenous variable, or control variable, *log(rincome)*. Right of the `|` shows the exogenous variable *log(rincome)* and instrumental variable *salestax*.

```{r}
cig_ivreg2 <- ivreg(log(packs) ~ log(rprice) + log(rincome) | log(rincome) + 
                    salestax, data = c1995)
summary(cig_ivreg2)
```

The estimated regression equation is
$$log(packs)=9.4307−1.1434log(rprice)+0.2145log(rincome)$$
A 1\% increase in cigarette price is associated with a 1.14\% decrease in cigarette sales. The impact is statisically significant at 1\% level.

Now explore a model with one control (income) and two instruments (sales tax and cigtax). When the number of instruments exceeds the number of endogenous variables, the model is overidentified. This allows us to test the validity of the instruments using an overidentification test. More instruments can also help reduce standard errors in 2SLS, improving the statistical power of estimates.

Same as the previous model, left of the `|` shows the dependent variable, endogenous variable, and the control variable. Right of the `|` shows the control variable, and **two** instrumental variables. 

```{r}
cig_ivreg3 <- ivreg(log(packs) ~ log(rprice) + log(rincome) | 
                    log(rincome) + salestax + cigtax, data = c1995)
summary(cig_ivreg3)
```

The estimated regression equation is
$$log(packs)=9.8950−1.2774log(rprice)+0.2804log(rincome)$$
A 1% increase in cigarette price is associated with a 1.28% decrease in cigarette sales. The effect is highly significant.

# Accessing Instrumental Validity

## Weak instrument test

To ensure the instruments (sales tax and cigarette tax) are strong predictors of the endogenous variable (log(rprice)), we perform a weak instrument test using the first-stage F-statistic.

Rule of thumb: If F > 10, the instruments are strong; if F < 10, they are weak, leading to biased IV estimates.

## Overidentification test (Sargan-Hansen)

When there are more instruments than endogenous variables, the model is overidentified, allowing us to test whether the instruments are truly exogenous. The Sargan-Hansen J-test checks if the instruments are uncorrelated with the error term. A **high** p-value (>0.05) suggests that the instruments are **valid**, while a low p-value indicates that at least one instrument may be invalid. This test helps ensure that our IV estimates are not biased due to instrument endogeneity.

## Endogeneity test (Wu-Hausman)

The Wu-Hausman test checks whether an explanatory variable is endogenous, meaning it is correlated with the error term. If endogeneity is present, OLS estimates are biased and inconsistent, making IV estimation necessary. The test compares the results of OLS and IV regressions—if the two estimates significantly differ, it suggests that the endogenous variable should be instrumented. A low p-value (p < 0.05) indicates that OLS is biased, confirming the need for IV, while a high p-value suggests that OLS may be sufficient. In R, this test can be performed using summary(iv_model, diagnostics = TRUE).

We can use one function to conduct all three tests:

```{r}
summary(cig_ivreg3,diagnostics = T)
```

Results show that:

First, for weak instrument test, F>10, confirming that sales tax and cigarette tax are strongly correelated with price. Weak instrument bias is not a concern.

Second, for endogeneity test, the null hypothesis is *log(rprice)* is exogenous, meaning we can use OLS model instead of instrumental variable. Since p>0.05, we fail to reject the null hypothesis, meaning we have evidence for endogeneity and instrumental variable is preferred. 

Finally, for overidentification test, the null hypothesis is the instruments are valid (uncorrelated with the error term). Since p>0.05, we fail to reject the null hypothesis, meaning both sales tax and cigarette tax are likely valid instruments.



